A Plasma Torus Around a Young Low-Mass Star

### 2025-05-25 paper/ms.tex
update

% upon AAS submission
\documentclass[11pt,twocolumn,tighten,linenumbers]{aastex7}
%\documentclass[11pt,twocolumn,tighten,linenumbers,trackchanges]{aastex7}
% drafting / arxiv
%\documentclass[11pt,twocolumn,modern]{aastex7}
%\turnoffedit

@@

\begin{abstract}
  A small fraction of red dwarfs younger than $\sim$100 million years
  show structured, periodic optical light curves suggestive of
  transiting opaque material that corotates with the star.  However, the
  composition, origin, and even the existence of this material are
  uncertain. The main alternative hypothesis is that these complex
  periodic variables (CPVs) are explained by complex distributions of
  bright or dark regions on the stellar surfaces.  Here, we present
  time-series spectroscopy and photometry of a rapidly-rotating
  ($P$=3.9\,hr) CPV, TIC\,141146667.  The spectra show sinusoidal
  time-varying H$\alpha$ emission at twice to four times the star's
  equatorial velocity, providing direct evidence for cool
  ($\lesssim$10$^4$\,K) plasma clumps trapped in corotation around a
  CPV.  These data support the idea that young, rapidly-rotating M
  dwarfs can sustain warped \added{tori of cool plasma, similar to
  other rapid magnetic rotators.}
  \added{Outstanding questions include whether dust clumps 
  in these plasma tori explain CPV light curves, and
  whether the tori originate from the star or are fed by external
  sources.}
  %It remains unclear whether these structures originate intrinsically
  %from the star or are fed by external sources. Likewise,
  %the mechanism by which their optically thick clumps arise is still
  %unknown.
  Rough estimates suggest $\gtrsim$10\% of M dwarfs host
  similar structures during their early lives.
\end{abstract}

@@

\citet{Townsend2005} outlined the physics of how rapidly-rotating stars
with strong magnetic fields can support plasma tori.  When the
magnetospheric radii $R_{\rm m}$ of such stars exceeds their Keplerian
corotation radii $R_{\rm c}$, the effective potential experienced by
charged particles \added{along any given field line }has a local minimum outside $R_{\rm c}$, which enables
charged material to accumulate in a torus
\citep{Petit2013,Daley-Yates2024}.  Warps can occur when there is
misalignment between the spin and magnetic axes.  In general, the
material in these centrifugally-supported magnetospheres need neither
transit nor be opaque in broadband optical light.

In a follow-up study, \citet{Townsend2008} synthesized light curves for
these systems, assuming that the optical depth scaled linearly with
density.  They found that W-shaped eclipses, similar to those seen for
TIC\,141146667, can occur when the spin and magnetic axes are moderately
(15-45$^\circ$) misaligned.  The movies associated with their work are
available
online\footnote{\url{http://user.astro.wisc.edu/\~townsend/static.php?ref=rrm-movies\#Download\_Bundles}
last accessed 23 April 2025.}; for a $\omega/\omega_{\rm
c}$$\approx$0.5, edge-on system like TIC\,141146667, the photometric
eclipses have the correct shape, and the H$\alpha$ emission similarly
exhibits double-peaked behavior.  The plasma in this model accumulates
at antipodes 180$^\circ$ apart because the deepest local minima in the
effective potential exist along the line of intersection between the
rotational and magnetic planes (see Equation~22 of
\citealt{Townsend2005}).

@@

Circumstellar H$\alpha$ emission might be sourced either from resonant
scattering of stellar H$\alpha$ photons, or from radiative
recombination.  We neglect scattering because Figure~\ref{fig:spec}
and Figure~\ref{fig:hirescuts} show the amplitude of the
circumstellar H$\alpha$ emission varying by a factor of $\approx$5, in
a manner uncorrelated with any variability in the chromospheric line
core.  The volume emissivity under case B recombination can be written
\begin{equation} j_{\rm H\alpha} = n_{\rm e} n_{\rm p} \alpha^{\rm
eff}_{\rm H\alpha} h \nu_{\rm H\alpha}, \end{equation} where $n_{\rm
e}$ and $n_{\rm p}$ are the electron and proton densities, and
$\alpha^{\rm eff}_{\rm H\alpha}$ is the effective recombination
coefficient, defined to include all recombination routes that produce
an H$\alpha$ photon.  For hydrogen with temperatures between
1,000-10,000\,K, $\alpha^{\rm eff}_{\rm H\alpha}$ is typically on the
order of $10^{-12}$ to $10^{-13}$\,cm$^3$\,s$^{-1}$
\citep{Hummer1987,Draine2011}.  Neglecting the effects of atoms other
than hydrogen, we can assume an ionization fraction $x$, such that
$n_{\rm e} = n_{\rm p} = x n_{\rm H}$, for $n_{\rm H}$ the hydrogen
number density.  Let $L_{\rm H\alpha} = j_{\rm H\alpha} V$, for $V$
the volume of the emitting hydrogen.  The luminosity of circumstellar
hydrogen emission, $L_{\rm H\alpha}$, is an observable: our SED
fitting routine yields $L_\star$$\approx$0.012\,$L_\odot$, which
implies that the stellar H$\alpha$ line radiates at
$\approx$1.0$\times$$10^{28}$erg\,s$^{-1}$.  The luminosity of the
clumps $L_{\rm H\alpha}$ are of order one tenth that of the star.  If
we approximate the emitting volume as a homogeneous sphere of radius
$r$, we can write
\begin{align}
  n_{\rm H} &= 
  1 \cdot 10^{11}\,{\rm cm}^{-3}
  \left(
    \frac{0.5}{x}
  \right)
  \left( 
    \frac{ L_{\rm H\alpha} }{ 10^{27}\,{\rm erg\,s^{-1}} }
    \frac{ 10^{-13}\,{\rm cm^{3}\,s^{-1}} }{ \alpha^{\rm eff}_{\rm H\alpha} }
  \right)^{1/2}
  \left(
    \frac{ 0.1\,R_\odot }{ r }
  \right)^{3/2}.
  \label{eq:numdensity}
\end{align}
For a uniform density clump, this suggests a total gas mass of $M_{\rm
gas}\approx 2\times10^{17}$\,g\added{, similar to those derived for
other cool stars \citep{VillarrealDAngelo2019}.}.  We emphasize that
Equation~\ref{eq:numdensity} in intended to provide only an order of
magnitude estimate for the number density implied by the observed
H$\alpha$ emission.  In detail, the effective recombination rate and
the ionization fraction each vary with density and temperature; a more
thorough estimate would iteratively solve the equations of detailed
balance and radiative transfer (e.g.~\citealt{CollierCameron1989}
Figure~8), and potentially also consider departures from local
thermodynamic equilibrium.

>>>>

% upon AAS submission
%\documentclass[11pt,twocolumn,tighten,linenumbers]{aastex7}
\documentclass[11pt,twocolumn,tighten,linenumbers,trackchanges]{aastex7}
% drafting / arxiv
%\documentclass[11pt,twocolumn,modern]{aastex7}
%\turnoffedit

@@

\begin{abstract}
  A small fraction of red dwarfs younger than $\sim$100 million years
  show structured, periodic optical light curves suggestive of
  transiting opaque material that corotates with the star.  However, the
  composition, origin, and even the existence of this material are
  uncertain. The main alternative hypothesis is that these complex
  periodic variables (CPVs) are explained by complex distributions of
  bright or dark regions on the stellar surfaces.  Here, we present
  time-series spectroscopy and photometry of a rapidly-rotating
  ($P$=3.9\,hr) CPV, TIC\,141146667.  The spectra show sinusoidal
  time-varying H$\alpha$ emission at twice to four times the star's
  equatorial velocity, providing direct evidence for cool
  ($\lesssim$10$^4$\,K) plasma clumps trapped in corotation around a
  CPV.  These data support the idea that young, rapidly-rotating M
  dwarfs can sustain warped \added{tori of cool plasma, similar to
  other rapidly-rotating magnetic stars.}
  \added{Outstanding questions include whether dust clumps 
  in these plasma tori explain CPV light curves, and
  whether the tori originate from the star or are fed by external
  sources.}
  %It remains unclear whether these structures originate intrinsically
  %from the star or are fed by external sources. Likewise,
  %the mechanism by which their optically thick clumps arise is still
  %unknown.
  Rough estimates suggest $\gtrsim$10\% of M dwarfs host
  similar structures during their early lives.
\end{abstract}

@@

\citet{Townsend2005} outlined the physics of how rapidly-rotating
stars with strong magnetic fields can support plasma tori \added{\citep[see
also][]{Nakajima1985,Ferreira2000,Petit2013,Daley-Yates2024}}.  When
the magnetospheric radii $R_{\rm m}$ of such stars exceeds their
Keplerian corotation radii $R_{\rm c}$, the effective potential
experienced by charged particles \added{along any given field line
}has a local minimum outside $R_{\rm c}$, which enables charged
material to accumulate in a torus.  Warps can occur when there is
misalignment between the spin and magnetic axes.  In general,
the\added{ regions} in these centrifugally-supported
magnetospheres\added{ with the densest plasma accumulations} need
neither transit nor be opaque in broadband optical light.

In a follow-up study, \citet{Townsend2008} synthesized light curves for
these systems, assuming that the optical depth scaled linearly with
plasma density.  They found that W-shaped eclipses, similar to those seen for
TIC\,141146667, can occur when the spin and magnetic axes are moderately
(15-45$^\circ$) misaligned.  The movies associated with their work are
available
online\footnote{\url{http://user.astro.wisc.edu/\~townsend/static.php?ref=rrm-movies\#Download\_Bundles}
last accessed 23 April 2025.}; for a $\omega/\omega_{\rm
c}$$\approx$0.5, edge-on system like TIC\,141146667, the photometric
eclipses have the correct shape, and the H$\alpha$ emission similarly
exhibits double-peaked behavior.  The plasma in this model accumulates
at antipodes 180$^\circ$ apart because the deepest local minima in the
effective potential exist along the line of intersection between the
rotational and magnetic planes (see Equation~22 of
\citealt{Townsend2005}).

@@

Circumstellar H$\alpha$ emission might be sourced either from resonant
scattering of stellar H$\alpha$ photons, or from radiative
recombination.  We neglect scattering because Figure~\ref{fig:spec}
and Figure~\ref{fig:hirescuts} show the amplitude of the
circumstellar H$\alpha$ emission varying by a factor of $\approx$5, in
a manner uncorrelated with any variability in the chromospheric line
core.  The volume emissivity under case B recombination can be written
\begin{equation} j_{\rm H\alpha} = n_{\rm e} n_{\rm p} \alpha^{\rm
eff}_{\rm H\alpha} h \nu_{\rm H\alpha}, \end{equation} where $n_{\rm
e}$ and $n_{\rm p}$ are the electron and proton densities, and
$\alpha^{\rm eff}_{\rm H\alpha}$ is the effective recombination
coefficient, defined to include all recombination routes that produce
an H$\alpha$ photon.  For hydrogen with temperatures between
1,000-10,000\,K, $\alpha^{\rm eff}_{\rm H\alpha}$ is typically on the
order of $10^{-12}$ to $10^{-13}$\,cm$^3$\,s$^{-1}$
\citep{Hummer1987,Draine2011}.  Neglecting the effects of atoms other
than hydrogen, we can assume an ionization fraction $x$, such that
$n_{\rm e} = n_{\rm p} = x n_{\rm H}$, for $n_{\rm H}$ the hydrogen
number density.  Let $L_{\rm H\alpha} = j_{\rm H\alpha} V$, for $V$
the volume of the emitting hydrogen.  The luminosity of circumstellar
hydrogen emission, $L_{\rm H\alpha}$, is an observable: our SED
fitting routine yields $L_\star$$\approx$0.012\,$L_\odot$, which
implies that the stellar H$\alpha$ line radiates at
$\approx$1.0$\times$$10^{28}$erg\,s$^{-1}$.  The luminosity of the
clumps $L_{\rm H\alpha}$ are of order one tenth that of the star.  If
we approximate the emitting volume as a homogeneous sphere of radius
$r$, we can write
\begin{align}
  n_{\rm H} &= 
  1 \cdot 10^{11}\,{\rm cm}^{-3}
  \left(
    \frac{0.5}{x}
  \right)
  \left( 
    \frac{ L_{\rm H\alpha} }{ 10^{27}\,{\rm erg\,s^{-1}} }
    \frac{ 10^{-13}\,{\rm cm^{3}\,s^{-1}} }{ \alpha^{\rm eff}_{\rm H\alpha} }
  \right)^{1/2}
  \left(
    \frac{ 0.1\,R_\odot }{ r }
  \right)^{3/2}.
  \label{eq:numdensity}
\end{align}
For a uniform density clump, this suggests a total gas mass of $M_{\rm
gas}\approx 2\times10^{17}$\,g\added{, which is similar to masses derived for
plasma clumps around other cool stars \citep{VillarrealDAngelo2019}.}.  We emphasize that
Equation~\ref{eq:numdensity} in intended to provide only an order of
magnitude estimate for the number density implied by the observed
H$\alpha$ emission.  In detail, the effective recombination rate and
the ionization fraction each vary with density and temperature; a more
thorough estimate would iteratively solve the equations of detailed
balance and radiative transfer (e.g.~Figure~8 of \citealt{CollierCameron1989}),
and potentially also consider departures from local thermodynamic equilibrium.
